---
title: "Joining Data in R with Dplyr"
output: html_notebook
---

#Joining Data in R with Dplyr

##Mutating Joins

left-join
right-join
inner-join
full-join

###A basic join
As Garrett mentioned in the video, left_join() is the basic join function in dplyr. You can use it whenever you want to augment a data frame with information from another data frame.

For example, left_join(x, y) joins y to x. The second dataset you specify is joined to the first dataset. Keep that in mind as you go through the course.

For this exercise, the code in the editor sets up a left join but doesn't finish it. It's up to you to complete it.

####A second join
100xp
You should always check the output of your joins. Just because a join worked, doesn't mean that it worked as you expected.

For example, the code in the editor joins the same two datasets that you joined in the previous exercise, but it returns a different result. Can you tell what went wrong?
```{r leftJoin}
library(tidyverse)
artists <- read_csv("./data/artists.csv")
bands <- read_csv("./data/bands.csv")

# Complete the code to join artists to bands
bands2 <- left_join(bands, artists, by = c("first", "last"))

# Examine the results
bands2

# Fix the code to recreate bands2
left_join(bands, artists, by = "first")
left_join(bands, artists, by = c("first", "last"))

```
###A right join
There is more than one way to execute a left join. Knowing multiple methods will make you a more versatile data scientist, especially as you try to fit joins into pipes created with %>%.

In this exercise, you'll recreate bands2 once more, but this time without using a left_join().
```{r rightJoin}
# Finish the code below to recreate bands3 with a right join
bands2 <- left_join(bands, artists, by = c("first", "last"))
bands3 <- right_join(artists, bands, by = c("first", "last"))

# Check that bands3 is equal to bands2
setequal(bands2, bands3)


```

###Inner joins and full joins
You may have noticed that some of the songs in songs correspond to some of the albums in albums. Suppose you want a new dataset that contains all of the songs for which you have data from both albums and songs. How would you make it?

The artists and bands datasets also share some information. What if you want to join these two datasets in such a way that you retain all of the information available in both tables, without throwing anything away?

You can think of inner joins as the most strict type of join: they only retain observations that appear in both datasets. In contrast, full joins are the most permissive type of join: they return all of the data that appears in both datasets (often resulting in many missing values).

Recall that, *_join(x, y) joins y to x. The second dataset you specify is joined to the first dataset.

```{r innerFullJoin}
albums <- read_csv("./data/albums.csv")
songs <- read_csv("./data/songs.csv")
# Join albums to songs using inner_join()
inner_join(songs, albums, by = "album")

# Join bands to artists using full_join()
full_join(artists, bands, by = c("first", "last"))

```

###Pipes
You can combine dplyr functions together with the pipe operator, %>%, to build up an analysis step-by-step. %>% takes the result of the code that comes before it and "pipes" it into the function that comes after it as the first argument of the function.

So for example, the two pieces of code below do the same thing:

full_join(artists, bands, 
          by = c("first", "last"))

artists %>% 
  full_join(bands, by = c("first", "last"))
Pipes are so efficient for multi-step analysis that you will use them for the remainder of the exercises in this course. (If you need a refresher on the pipe operator, check out Data Manipulation in R with dplyr course!)
```{r pipes}
# Find guitarists in bands dataset (don't change)
temp <- left_join(bands, artists, by = c("first", "last"))
temp <- filter(temp, instrument == "Guitar")
select(temp, first, last, band)

# Reproduce code above using pipes
bands %>% 
  left_join(artists, by = c("first", "last")) %>%
  filter(instrument == "Guitar") %>%
  select(first, last, band)


goal <- as_tibble(read_csv("./data/goal.csv"))
# Examine the contents of the goal dataset
goal

# Create goal2 using full_join() and inner_join() 
goal2 <- artists %>% 
        full_join(bands, by = c("first", "last")) %>%
        inner_join(songs, by = c("first", "last"))
        
str(goal2)  
  
# Check that goal and goal2 are the same
setequal(goal, goal2)

# Create one table that combines all information
artists %>% 
        full_join(bands, by = c("first", "last")) %>%
        full_join(songs, by = c("first", "last")) %>%
        full_join(albums, by = c("album","band"))
```

##Filtering Joins and Set Operators

###Apply a semi-join
As you saw in the video, semi-joins provide a concise way to filter data from the first dataset based on information in a second dataset.

For example, the code in the editor uses semi_join() to create a data frame of the artists in artists who have written a song in songs.
```{r semiJoin}
library(tidyverse)
# View the output of semi_join()
artists %>% 
  semi_join(songs, by = c("first", "last"))

# Create the same result
artists %>% 
  right_join(songs, by = c("first", "last")) %>% 
  filter(!is.na(instrument)) %>% 
  select(first, last, instrument)

```
###Exploring with semi-joins

Semi-joins provide a useful way to explore the connections between multiple tables of data.

For example, you can use a semi-join to determine the number of albums in the albums dataset that were made by a band in the bands dataset.
```{r exploreSemiJoin}

albums %>% 
        semi_join(bands, by = "band") %>% 
        nrow()


```
###We've attempted to rewrite this semi-join as a filter. Will it return the same results?
```{r}
tracks <- read_csv("./data/tracks.csv")
matches <- read_csv("./data/matches.csv")
tracks %>% semi_join(
  matches,
  by = c("band", "year", "first")
)

tracks %>% filter(
  (band == "The Beatles" & 
     year == 1964 & first == "Paul") |
    (band == "The Beatles" & 
       year == 1965 & first == "John") |
    (band == "Simon and Garfunkel" & 
       year == 1966 & first == "Paul")
)
```
###Apply an anti-join
As Garrett mentioned in the video, anti-joins provide a useful way to reason about how a mutating join will work before you apply the join.

For example, you can use an anti-join to see which rows will not be matched to a second dataset by a join.

Apply another anti-join
Anti-joins with anti_join() also provide a great way to diagnose joins that go wrong.

For example, they can help you zero-in on rows that have capitalization or spelling errors in the keys. These things will make your primary and secondary keys appear different to R, even though you know they refer to the same thing.
```{r antiJoin}

artists %>% 
        anti_join(bands, by = c("first", "last"))

labels <- read_csv("./data/labels.csv")

labels %>% 
        anti_join(albums, by = "album")
```
###Which filtering join?
Think you have filtering joins down? Let's check.

Which filtering join would you use to determine how many rows in songs match a label in labels?
```{r whichFilterJoin}
# Determine which key joins labels and songs
labels
songs

# Check your understanding
songs %>% 
  # Find the rows of songs that match a row in labels
  semi_join(labels, by = "album") %>% 
  # Number of matches between labels and songs
  nrow()
```
###Set Operations
union - all of both datasets
intersect - the shared observations from each data set
setdiff - the observations from one dataset that are not in the other

How many songs are there?
We have loaded two datasets in your workspace, aerosmith and greatest_hits, each of which represents an album from the band Aerosmith. Each row in either of the datasets is a song on that album.

How many unique songs do these two albums contain in total?
```{r setOperations}
aerosmith <- read_csv("./data/aerosmith/aerosmith.csv")
greatest_hits <- read_csv("./data/aerosmith/greatest_hits.csv")

#all songs from aerosmith and greatest_hits
aerosmith %>% 
  # Create the new dataset using a set operation
  union(greatest_hits, by = "song") %>% 
  # Count the total number of songs
  nrow()

#songs from aerosmith on greatest hits
aerosmith %>% 
        intersect(greatest_hits, by = "song")

```
###Live! Bootleg songs
live contains the songs from a third Aerosmith album, Live! Bootleg.

Which songs are on Live! Bootleg but not on Greatest Hits? Notice that the length of songs may be different when they are performed live.
```{r moreSetOperations}
live <- read_csv("./data/aerosmith/live.csv")

# Select the song names from live
live_songs <- live %>% select(song)

# Select the song names from greatest_hits
greatest_songs <- greatest_hits %>% select(song)

# Create the new dataset using a set operation
live_songs %>% 
  setdiff(greatest_songs)

```
###Multiple operations
There is no set operation to find rows that appear in one data frame or another, but not both. However, you can accomplish this by combining set operators, as Garrett showed in the video.

Can you think of a combination that would answer the question, "Which songs appear on one of Live! Bootleg or Greatest Hits, but not both?"
```{r multipleSetOperations}
# Select songs from live and greatest_hits
# Select the song names from live
live_songs <- live %>% select(song)
greatest_songs <- greatest_hits %>% select(song)

# Return the songs that only exist in one dataset
live_only <- live_songs %>% 
        setdiff(greatest_songs)
great_only <- greatest_songs %>% 
        setdiff(live_songs)
liveOnly_great_only <- live_only %>% 
        union(great_only)

# Select songs from live and greatest_hits
live_songs <- live %>% select(song)
greatest_songs <- greatest_hits %>% select(song)

# Return the songs that only exist in one dataset
all_songs <- live_songs %>% union(greatest_songs)
common_songs <- live_songs %>% intersect(greatest_songs)
courseSolution  <-  all_songs %>% setdiff(common_songs)
setequal(liveOnly_great_only, courseSolution)
```
###Apply setequal
definitive and complete contain the songs that appear in competing Led Zeppelin anthologies: The Definitive Collection and The Complete Studio Recordings, respectively.

Both anthologies claim to contain the complete studio recordings of Led Zeppelin, but do the anthologies contain the same exact songs?

Apply setequal again
A few exercises ago, you saw that an intersect() is analagous to a semi_join() when two datasets contain the same variables and each variable is used in the key.

Under these conditions, setdiff() is also analagous to one of the filtering joins.
```{r setEqual}
definitive <- read_csv("./data/definitive.csv")
complete <- read_csv("./data/complete.csv")

# Check if same order: definitive and complete
identical(definitive, complete)

# Check if any order: definitive and complete
setequal(definitive, complete)

# Songs in definitive but not complete
definitive %>% 
        setdiff(complete)


# Songs in complete but not definitive
complete %>%
        setdiff(definitive)


# Return songs in definitive that are not in complete
definitive %>% 
  anti_join(complete, by = c("song", "album"))

# Return songs in complete that are not in definitive
complete %>% 
  anti_join(definitive, by = c("song", "album"))

```
###Comparing albums
It appears that The Definitive Collection contains songs from the soundtrack of The Song Remains the Same, a movie filmed during a live Led Zeppelin concert. Is this the only difference between The Definitive Collection and The Complete Studio Recordings?
```{r compareDatasets}
soundtrack <- read_csv("./data/soundtrack.csv")

# Check if same order: definitive and union of complete and soundtrack
identical(definitive, union(complete, soundtrack, by = c("song", "album")))



# Check if any order: definitive and union of complete and soundtrack
setequal(definitive, union(complete, soundtrack, by = c("song", "album")))


```
##Assembling Data
Binds
bind_row
bind_cols
###Which bind?
side_one and side_two contain tracks from sides one and two, respectively, of Pink Floyd's famous album The Dark Side of the Moon.

Bind these datasets into a single table using a dplyr function. Which type of bind should you use?
```{r binds}
library(tidyverse)
side_one <- as_tibble(read_csv("./data/side_one.csv"))
side_two <- as_tibble(read_csv("./data/side_two.csv"))

# Examine side_one and side_two
side_one
side_two

# Bind side_one and side_two into a single dataset
side_one %>% 
  bind_rows(side_two)

```
###Bind rows
discography and jimi contain all of the information you need to create an anthology dataset for the band The Jimi Hendrix Experience.

discography contains a data frame of each album by The Jimi Hendrix Experience and the year of the album.

jimi contains a list of data frames of album tracks, one for each album released by The Jimi Hendrix Experience. As Garrett explained in the video, you can pass bind_rows() a list of data frames like jimi to bind together into a single data frame.
```{r bindRows}
discography <- as_tibble(read_csv("./data/jimi_hendrix/discography.csv"))
jimi1 <- as_tibble(read_csv("./data/jimi_hendrix/jimi1.csv"))
jimi2 <- as_tibble(read_csv("./data/jimi_hendrix/jimi2.csv"))
jimi3 <- as_tibble(read_csv("./data/jimi_hendrix/jimi3.csv"))
bind_rows(jimi1, jimi2, jimi3)
# Examine discography and jimi
discography
jimi

  bind_rows('Are You Experienced' = jimi1,
            'Axis: Bold As Love' = jimi2,
            'Electric Ladyland' = jimi3,
            .id = "album") %>% 
  # Make a complete data frame
  left_join(discography, by = "album")

```
###Bind Columns
Let's make a compilation of Hank Williams' 67 singles. To do this, you can use hank_years and hank_charts:

hank_years contains the name and release year of each of Hank Williams' 67 singles.
hank_charts contains the name of each of Hank Williams' 67 singles as well as the highest position it earned on the Billboard sales charts.
Each dataset contains the same songs, but hank_years is arranged chronologically by year, while hank_charts is arranged alphabetically by song title.
```{r bindColumns}
hank_years <- as_tibble(read_csv("./data/hank_williams/hank_years.csv"))
hank_charts <- as_tibble(read_csv("./data/hank_williams/hank_charts.csv"))
# Examine hank_years and hank_charts
hank_years
hank_charts

hank_years %>% 
  # Reorder hank_years alphabetically by song title
  arrange(song) %>% 
  # Select just the year column
  select(year) %>% 
  # Bind the year column
  bind_cols(hank_charts) %>% 
  # Arrange the finished dataset
  arrange(year, song)

?arrange
```

